Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-29 paper/main.tex
minor

For both feasibility and the frequent validation of cross-species generalization during
development, we
selected only some of the remaining genomes for direct training.
The selection of training genomes also requires balancing multiple conflicting 
tradeoffs in general. On the one hand, we want to train with as much data as possible, but 
on the other
not all data has sufficient quality to enable a powerful generalization. It is 
also very desirable to have diverse training genomes with a broad phylogenetic spread, 
a variety of genome sizes and average gene lengths as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, it also may be difficult 
for the model learn and generalize if the genomic patterns inside the data are 
too different from each other. Practically, it was important for us to limit the size of
our training genomes set to be small enough that enough that we could get experimental 
results within a couple of days and thereby be able to test many different data and 
model configurations.

@@

For the training itself, we divide each continous genomic sequence into 20,000 bp 
long subsequences, for which one-hot vectors of the base pairs and annotations 
are generated and respectively used as input and label for the neural network 
together with the sample weights. We append zero padding if the sequences (or
sequence ends) are shorter than 20,000 bp. 
If a subsequence is fully marked as erroneous, we exclude it from all analyses.

@@

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (about 400 genes and several days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by AUGUSTUS. An untrained model was setup with ``new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with ``optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by AUGUSTUS were imported into GeenuFF and exported as HDF5
files in the same manor as the reference, allowing for the direct comparison with both the reference and 
Helixer predictions. While tenable for some 237 species, this method allowed neither training
nor prediction of UTR regions with AUGUSTUS, so the metric Subgenic F1, which disregards the class UTR,
was used for all comparisons between AUGUSTUS and Helixer (see Section \ref{sec:metrics} for more details) .

@@

Helixer models do not yet have postprocessing
to make finalized single predictions, but instead output basewise probabilities.
We see that the model exibits higher uncertainty around transitions between annotation 
classes, for instance between UTR and CDS, or more dramatically between UTR and 
intergenic even where Helixer predictions closely match the reference and RNAseq data (a). 
Helixer's uncertainty around UTR <-> intergenic transitions may relate to a
fundamentally harder problem (there is no conserved motif at the site as is observed
for splice sites and start/stop codons), lack of a precise one 
base pair biological site \citep{carninci2006genome,hon2013quantification}
or noise in the reference, which we observed relative to the RNAseq data 
(Figures S\ref{supfig:UTRs_animals}-\ref{supfig:UTRs_plants}). % supplemental UTR figures, 1/species

Helixer models also sometimes showed uncertainty for larger regions.
In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (Figures \ref{fig:cov_example_main}b-c,S\ref{supfig:cov_example_02}c). However, in the extreme,
there are cases where the Helixer model exibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (Figures S\ref{supfig:cov_example_02}c, S\ref{supfig:cov_example_02}c-d). 
Notably, in one of the examples (S\ref{supfig:cov_example_02}d))
the RNAseq shows evidence of alternative splicing; 
and in another (S\ref{supfig:cov_example_02}c) Helixer's
prediction falls between that of the reference and AUGUSTUS.

@@

Coverage and Spliced coverage were broken down by the confusion matrix (of both
Helixer vs AUGUSTUS and Helixer vs the reference) for all genomes (Figure \ref{fig:average_vs_augustus}). % H vs A main, H vs R supp
Where tools were in agreement (far left), coverage and spliced coverage closely matched
expectations. Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage. In all conflicts the amount of RNAseq support
fell between the cases where tools agreed, indicating that all options were at least
capable of finding weak-spots in the other annotations.
In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of AUGUSTUS. Specifically, for base pairs where one tool said CDS and
the other intergenic or intron, there was more coverage when 
Helixer said CDS (Figure \ref{fig:average_vs_augustus}a). Similarly, for base pairs where one tool said intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer said intron (Figure \ref{fig:average_vs_augustus}b). 
These patterns were consistent in direction but varied in magnitude in the individual species,
with the expception of {\it P. marinus} where Helixer and AUGUSTUS performed comparably
at differentiating CDS and introns (Figure S\ref{supfig:each_sp_vs_augustus}).

@@

\section{Discussion}
We indtroduce Helixer a novel, deep-learning based tool for gene annotation
prediction. Helixer outperforms AUGUSTUS on base pair wise metrics and on 
consistency with independent RNASeq data.

>>>>

For the selection of training genomes we had to balance multiple conflicting 
tradeoffs. On the one hand, we want to train with as much data as possible, but 
on the other
not all data has sufficient quality to enable a powerful generalization. It is 
also very desirable to have diverse training genomes with a broad phylogenetic spread, 
a variety of genome sizes and average gene lengths as the model is increasingly unlikely to generalize 
well beyond the borders given by the training data. However, it also may be difficult 
for the model learn and generalize if the genomic patterns inside the data are 
too different from each other. Practically, it was important for us to limit the size of
our training genomes set to be small enough that enough that we could get experimental 
results within a couple of days and thereby be able to test many different data and 
model configurations.

@@

For the training itself, we divide each continous genomic sequence into 20,000 bp 
long subsequences, for which one-hot vectors of the base pairs and annotations 
are generated and respectively used as input and label for the neural network 
together with the sample weights. We append zero padding if the subsequences  
are shorter than 20,000 bp. 
If a subsequence is fully marked as erroneous, we exclude it from all analyses.

@@

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (about 400 genes and several days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by AUGUSTUS. An untrained model was setup with ``new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with ``optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by AUGUSTUS were imported into GeenuFF and exported as HDF5
files in the same manor as the reference, allowing for the direct comparison with both the reference and 
Helixer predictions. While still feasible for some 237 species, this method allowed neither training
nor prediction of UTR regions with AUGUSTUS, so the metric Subgenic F1, which disregards the class UTR,
was used for all comparisons between AUGUSTUS and Helixer (see Section \ref{sec:metrics} for more details) .

@@

Helixer models do not yet have postprocessing
to make finalized single predictions, but instead output basewise probabilities.
We see that the model exibits higher uncertainty around transitions between annotation 
classes, for instance between UTR and CDS, or more dramatically between UTR and 
intergenic even where Helixer predictions closely match the reference and RNAseq data (a). 
Helixer's uncertainty around transitions from UTR to intergenic regions may relate to a
fundamentally harder problem (there is no conserved motif at the site as is observed
for splice sites and start/stop codons), lack of a precise one 
base pair biological site \citep{carninci2006genome,hon2013quantification}
or noise in the reference, which we observed relative to the RNAseq data 
(Figures S\ref{supfig:UTRs_animals}-\ref{supfig:UTRs_plants}). % supplemental UTR figures, 1/species

Helixer models also sometimes showed uncertainty for larger regions.
In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (Figures \ref{fig:cov_example_main}b-c,S\ref{supfig:cov_example_02}c). However, in the extreme,
there are cases where the Helixer model exibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (Figures S\ref{supfig:cov_example_02}c-d). 
Notably, in one of the examples (Figure S\ref{supfig:cov_example_02}d)
the RNAseq shows evidence of alternative splicing; 
and in another (Figure S\ref{supfig:cov_example_02}c) Helixer's
prediction falls between that of the reference and of AUGUSTUS.

@@

Coverage and Spliced coverage were broken down by the confusion matrix of both
Helixer vs AUGUSTUS as well as Helixer vs the reference for all genomes (Figure \ref{fig:average_vs_augustus}). % H vs A main, H vs R supp
Where both tools agreee on far left in the figure, coverage and spliced coverage closely matched expectations. 
Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage. In all conflicts the amount of RNAseq support
fell between the cases where tools agreed, indicating that all options were at least
capable of finding weak-spots in the other annotations.

In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of AUGUSTUS. Specifically, for base pairs where one tool predicted CDS and
the other intergenic or intron, there was more coverage when 
Helixer predicted CDS (Figure \ref{fig:average_vs_augustus}a). Similarly, for base pairs where one tool predicted intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer predicted intron (Figure \ref{fig:average_vs_augustus}b). 
These patterns were consistent in direction but varied in magnitude in the individual species,
with the expception of {\it P. marinus} where Helixer and AUGUSTUS performed comparably
at differentiating between CDS and intron regions (Figure S\ref{supfig:each_sp_vs_augustus}).

@@

\section{Discussion}
With Helixer, we introduce a novel, deep-learning based tool for gene annotation
prediction. Helixer outperforms AUGUSTUS on base pair wise metrics and on 
consistency with independent RNASeq data.
